---
title: "Single sample gene set enrichment analysis"
author: "Menna Arafat"
date:  "2024-8-25"
output: html_document
Main_tutorial: https://www.bioconductor.org/packages/release/bioc/vignettes/GSVA/inst/doc/GSVA.html#:~:text=Gene%20set%20variation%20analysis%20(GSVA,from%20genes%20to%20gene%20sets.
---



## create output directory

```{r}

dir.create("output")
```

## BiocManager::install("fields")

```{r}
 suppressPackageStartupMessages({ 
library("GSVAdata")
library(MOFA2)   
library("GSVA")
library(GSEABase)
library(GSVAdata)
library(fgsea)
library(tibble)
library("RColorBrewer")
library("circlize")
library(ComplexHeatmap)
library(biomaRt)
library(dplyr)
library(plyr)
library(limma)
library(fields)
 })
```

## load your gmt as gene set (list of vectors)

```{r}

gset = gmtPathways("data/Human_GO_AllPathways_main.gmt.txt")
```

## filter gene set to have at least 5 terms pr set

```{r}
gset_sub= gset[lapply(gset, length) >= 5]

gset_main= gset_sub[!grepl("PMC",names(gset_sub))]

```

##  convert list to long formats for clusterprofiler

```{r}
gmt_long= stack(gset_sub) 
gmt_long= gmt_long[,c(2,1)] 
```

## load data

```{r}
list.files()
data= read.csv("data/Final.DataSet.csv") 
metadata= data.frame(sample= colnames(data)[-1],
                     condition= as.character(data[1,-1]))
data= data[-1,]
```

## change type

```{r}
data[,-1]= lapply(data[,-1], function(x) as.numeric(as.character(x))) 
sum(is.na(data))
str(data)
```

## id mapping

```{r}
list.files()
mapping= read.delim("data/idmapping.tsv" )
data$id= mapping$To[match(data$Sample, mapping$From)]
length(unique(data$id))
```

## remove duplicate ids

```{r}
data= data %>% mutate(mean= rowSums(dplyr::select(.,where(is.numeric)))/ncol(data)-2) %>% 
                         group_by(id) %>%
                         dplyr::slice(which.max(mean)) %>% 
                         ungroup() %>% dplyr::select(-c(mean, Sample)) %>% 
                         filter(!is.na(id)) %>% as.data.frame()
row.names(data)= data$id
data= data[,!grepl("id", colnames(data))]
head(data)
```

## determine the distribution of data usually gaussian for log transformed data, and poisson for FPKM count data

```{r}
general_mean= apply(data, 1, mean)
hist(general_mean, main = "Histogram of Data", xlab = "Values", breaks = 30)
```

## check whether it follows gaussian distribution 

```{r}
qqnorm(general_mean)
```

## check normality

```{r}
shapiro.test(general_mean)
```

## log transform

```{r}
data.log= log(data, base=2)
```

## check normality

```{r}
general_mean= apply(data.log, 1, mean)
shapiro.test(general_mean)
```


## check whether it follows poisson distribution 

```{r}
general_mean =apply(data.log, 1, mean) #' ## general_mean=lambda
poisson_quantiles = qpois(ppoints(length(general_mean)), lambda = general_mean)  #' ##  Theoretical quantiles, #' ## propability density function of poisson model probability distribution that models the number of times an event happens in a fixed interval of time or space given the average rate is constant, quantile function is essentially the inverse of this CDF, as it finds the k events corresponding to a specific cumulative probability
```

## plot the empirical quantiles against the theoretical quantiles.

```{r}
qqplot(poisson_quantiles, sort(general_mean), 
       main = "Q-Q Plot for Poisson Distribution",
       xlab = "Theoretical Quantiles (Poisson)",
       ylab = "Empirical Quantiles (Data)")
```

## run ssgsea
### we could set the distribution of the data throught the parameter kcdf = c("auto", "Gaussian", "Poisson"), for kcdf="auto", this will make GSVA to decide automatically the distribution

```{r}
gsva_obj = gsvaParam(as.matrix(data),
                      gset_sub, minSize=1, maxSize=500,
                      kcdf="auto")
ES = gsva(gsva_obj)
write.csv(ES, "output/single_sample_gsea_enrichment_scores.csv", row.names = T)
```
## heatmap
```{r}

shapiro.test(as.numeric(ES[1, ]))  #' ## p-value <= 0.05 indicative of non normal distribution 
shapiro.test(as.numeric(ES[, 1])) 

colorLegend <- c("darkolivegreen" ,"#ADFF2F" )
names(colorLegend) <- unique(metadata$condition)
sample.color.map = colorLegend[metadata$condition]
names(sample.color.map) <- metadata$sample

sampleClustering <- hclust(as.dist(1-cor(as.matrix(ES), method="spearman")), #' ## pearson #' ## spearman
                           method="complete")
geneSetClustering <- hclust(as.dist(1-cor(t(ES), method="spearman")),
                            method="complete")
palette <- colorRampPalette(c( "lightyellow2" ,"#DFC27D" , "#FCAA0FFF", "darkred"))(256)

png("output/heatmap_pathsway_signature_ssgsea.png", height=4100, width = 2650, res= 600)
heatmap(as.matrix(ES), ColSideColors=sample.color.map, xlab="samples",
        ylab="", margins=c(2, 20),
        col = palette ,
        labRow=substr(gsub("_", " ", gsub("^KEGG_|^REACTOME_|^BIOCARTA_", "", rownames(ES))), 1, 35),
        labCol="", scale="none", Colv=as.dendrogram(sampleClustering),
        Rowv=as.dendrogram(geneSetClustering)
         )
legend("right", inset = 0.001, names(colorLegend), fill=colorLegend, bg="white",
       cex = .6, title= "Phases") #' ##  #' ## x = .01, y =9
image.plot(zlim = range(ES, na.rm = TRUE),
           legend.only = TRUE, 
           horizontal = TRUE, 
           legend.shrink = 0.3, 
           legend.width = 0.9,
           legend.mar= 4.5,
           col = palette , 
           legend.position = c(0.5, 0.5), 
           inset = c(-1.1, 0),
           legend.args = list(text = "Enrichment Score (ES)", side = 3, line = .5, cex = 0.6))
dev.off()
```
